quollr: An R Package for Visualizing 2-D Models from Non-linear Dimension Reductions in High Dimensional Space

### 2025-05-15 paper-quollr.Rmd
Recomputed the RMSE for limb data

Non-linear dimension reduction (NLDR) techniques, such as t-distributed stochastic neighbor embedding (tSNE) [@laurens2008], uniform manifold approximation and projection (UMAP) [@leland2018], potential of heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm [@moon2019], large-scale dimensionality reduction Using triplets (TriMAP) [@amid2019], and pairwise controlled manifold approximation (PaCMAP) [@yingfan2021], create wildly different representations depending on the selected method and (hyper)parameter choices (Figure \@ref(fig:nldr-layouts)). It is difficult to determine whether any of these representations are accurate, which one is the best, or whether they have missed important structures.

@@

```{r, echo=TRUE}
fit_highd_model(
  highd_data = scurve, 
  nldr_data = scurve_umap, 
  bin1 = 15, 
  q = 0.1, 
  benchmark_highdens = 5)
```

@@

Limb muscle cells of mice in @tabula2018 are examined. There are $1067$ single cells, with $14997$ gene expressions. Following their pre-processing, tSNE was performed using ten principal components. \@ref(fig:nldr-layouts) (b) is the reproduction of the published plot. The question is whether this accurately represents the cluster structure in the data. Our method can help here, and also help to provide a better $2-\text{D}$ layout, as needed.

@@

error_limb <- error_limb |>
  mutate(a1 = round(a1, 2)) |>
  filter(bin1 >= 5) |>
  filter(a1 >= 0.03) |>
  group_by(method, a1) |>
  filter(MSE == min(MSE)) |>
  ungroup()

@@

error_plot_limb <- plot_mse(error_limb) +
  scale_x_continuous(breaks = sort(unique(error_limb$a1))[c(1, 5, 9, 13, 17, 21, 26)]) +
  scale_color_manual(values=c('#999999','#a65628','#e41a1c','#377eb8','#4daf4a','#ff7f00','#984ea3','#f781bf'))

@@

```{r limb-rmse, fig.cap="Assessing which of the 6 NLDR layouts on the limb muscle data  (shown in \\@ref(fig:nldr-layouts)) is the better representation using RMSE for varying binwidth ($a_1$). Colour  used for the lines and points in the left plot and in the scatterplots represents NLDR layout (a-f). Layout d is perform well at large binwidth (where the binwidth is not enough to capture the data struture) and poorly as bin width decreases. Layout f is the best choice.", out.width="100%"}

@@

```{r, eval=knitr::is_html_output()}
#| label: langevitour-limb-tsne-author-proj

@@

```{r, eval=knitr::is_html_output()}
#| label: langevitour-limb-tsne-best-proj

@@

```{r model-limb, fig.cap="Compare the published $2-\\text{D}$ layout (\\@ref(fig:nldr-layouts) b) and the $2-\\text{D}$ layout selected (\\@ref(fig:nldr-layouts) f) by RMSE plot (\\@ref(fig:limb-rmse)) from the tSNE, UMAP, PHATE, TriMAP, and PaCMAP with different (hyper)parameters. The Limb muscle data ($n =  1067$) has seven close different shaped clusters in $10\\text{-}D$.", out.width="100%", fig.width=15, fig.height=10}

>>>>

Non-linear dimension reduction (NLDR) techniques, such as t-distributed stochastic neighbor embedding (tSNE) [@laurens2008], uniform manifold approximation and projection (UMAP) [@leland2018], potential of heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm [@moon2019], large-scale dimensionality reduction Using triplets (TriMAP) [@amid2019], and pairwise controlled manifold approximation (PaCMAP) [@yingfan2021], create wildly different representations depending on the selected method and (hyper)parameter choices. It is difficult to determine whether any of these representations are accurate, which one is the best, or whether they have missed important structures.

@@

```{r, echo=TRUE, eval=FALSE}
fit_highd_model(
  highd_data = scurve, 
  nldr_data = scurve_umap, 
  bin1 = 15, 
  q = 0.1, 
  benchmark_highdens = 5)
```

@@

Limb muscle cells of mice in @tabula2018 are examined. There are $1067$ single cells, with $14997$ gene expressions. Following their pre-processing, tSNE was performed using ten principal components. Figure \@ref(fig:limb-rmse) (b) is the reproduction of the published plot. The question is whether this accurately represents the cluster structure in the data. Our method can help here, and also help to provide a better $2-\text{D}$ layout, as needed.

@@

error_limb <- error_limb |>
  mutate(a1 = round(a1, 2)) |>
  filter(bin1 >= 5) |>
  filter(a1 >= 0.03) |>
  group_by(method, a1) |>
  filter(RMSE == min(RMSE)) |>
  ungroup()

@@

error_plot_limb <- plot_rmse(error_limb) +
  scale_x_continuous(breaks = sort(unique(error_limb$a1))[c(1, 5, 9, 13, 17, 21, 26)]) +
  scale_color_manual(values=c('#999999','#a65628','#e41a1c','#377eb8','#4daf4a','#ff7f00','#984ea3','#f781bf'))

@@

```{r limb-rmse, fig.cap="Assessing which of the 6 NLDR layouts on the limb muscle data is the better representation using RMSE for varying binwidth ($a_1$). Colour  used for the lines and points in the left plot and in the scatterplots represents NLDR layout (a-f). Layout d is perform well at large binwidth (where the binwidth is not enough to capture the data struture) and poorly as bin width decreases. Layout f is the best choice.", out.width="100%"}

@@

```{r langevitour-limb-tsne-author-proj, eval=knitr::is_html_output()}

@@

```{r langevitour-limb-tsne-best-proj, eval=knitr::is_html_output()}

@@

```{r model-limb, fig.cap="Compare the published $2-\\text{D}$ layout (Figure \\@ref(fig:limb-rmse) b) and the $2-\\text{D}$ layout selected (Figure \\@ref(fig:limb-rmse) f) by RMSE plot (Figure \\@ref(fig:limb-rmse)) from the tSNE, UMAP, PHATE, TriMAP, and PaCMAP with different (hyper)parameters. The Limb muscle data ($n =  1067$) has seven close different shaped clusters in $10\\text{-}D$.", out.width="100%", fig.width=15, fig.height=10}

### 2025-05-15 paper-quollr.tex
Recomputed the RMSE for limb data

Non-linear dimension reduction (NLDR) techniques, such as t-distributed stochastic neighbor embedding (tSNE) \citep{laurens2008}, uniform manifold approximation and projection (UMAP) \citep{leland2018}, potential of heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm \citep{moon2019}, large-scale dimensionality reduction Using triplets (TriMAP) \citep{amid2019}, and pairwise controlled manifold approximation (PaCMAP) \citep{yingfan2021}, create wildly different representations depending on the selected method and (hyper)parameter choices (Figure \ref{fig:nldr-layouts}). It is difficult to determine whether any of these representations are accurate, which one is the best, or whether they have missed important structures.

@@

\begin{verbatim}
#> $nldr_obj
#> $nldr_obj$scaled_nldr
#> # A tibble: 5,000 x 3
#>     emb1  emb2    ID
#>    <dbl> <dbl> <int>
#>  1 0.707 0.839     1
#>  2 0.231 0.401     2
#>  3 0.232 0.215     3
#>  4 0.790 0.564     4
#>  5 0.761 0.551     5
#>  6 0.445 0.721     6
#>  7 0.900 0.137     7
#>  8 0.247 0.392     8
#>  9 0.325 0.542     9
#> 10 0.278 0.231    10
#> # i 4,990 more rows
#> 
#> $nldr_obj$lim1
#> [1] -14.42166  13.32655
#> 
#> $nldr_obj$lim2
#> [1] -12.43687  12.32455
#> 
#> 
#> $hb_obj
#> $a1
#> [1] 0.08326136
#> 
#> $a2
#> [1] 0.07210645
#> 
#> $bins
#> [1] 15 16
#> 
#> $start_point
#> [1] -0.10000000 -0.08923607
#> 
#> $centroids
#> # A tibble: 240 x 3
#>    hexID     c_x     c_y
#>    <int>   <dbl>   <dbl>
#>  1     1 -0.1    -0.0892
#>  2     2 -0.0167 -0.0892
#>  3     3  0.0665 -0.0892
#>  4     4  0.150  -0.0892
#>  5     5  0.233  -0.0892
#>  6     6  0.316  -0.0892
#>  7     7  0.400  -0.0892
#>  8     8  0.483  -0.0892
#>  9     9  0.566  -0.0892
#> 10    10  0.649  -0.0892
#> # i 230 more rows
#> 
#> $hex_poly
#> # A tibble: 1,440 x 3
#>    hex_poly_id       x       y
#>          <int>   <dbl>   <dbl>
#>  1           1 -0.1    -0.0412
#>  2           1 -0.142  -0.0652
#>  3           1 -0.142  -0.113 
#>  4           1 -0.1    -0.137 
#>  5           1 -0.0584 -0.113 
#>  6           1 -0.0584 -0.0652
#>  7           2 -0.0167 -0.0412
#>  8           2 -0.0584 -0.0652
#>  9           2 -0.0584 -0.113 
#> 10           2 -0.0167 -0.137 
#> # i 1,430 more rows
#> 
#> $data_hb_id
#> # A tibble: 5,000 x 4
#>     emb1  emb2    ID hexID
#>    <dbl> <dbl> <int> <int>
#>  1 0.707 0.839     1   205
#>  2 0.231 0.401     2   109
#>  3 0.232 0.215     3    65
#>  4 0.790 0.564     4   146
#>  5 0.761 0.551     5   146
#>  6 0.445 0.721     6   172
#>  7 0.900 0.137     7    58
#>  8 0.247 0.392     8   110
#>  9 0.325 0.542     9   141
#> 10 0.278 0.231    10    80
#> # i 4,990 more rows
#> 
#> $std_cts
#> # A tibble: 142 x 3
#>    hexID bin_counts std_counts
#>    <int>      <int>      <dbl>
#>  1    21         12     0.16  
#>  2    22         17     0.227 
#>  3    23         22     0.293 
#>  4    24         10     0.133 
#>  5    25          7     0.0933
#>  6    26         12     0.16  
#>  7    27          1     0.0133
#>  8    36         42     0.56  
#>  9    37         42     0.56  
#> 10    38         44     0.587 
#> # i 132 more rows
#> 
#> $tot_bins
#> [1] 240
#> 
#> $non_bins
#> [1] 142
#> 
#> $pts_bins
#> # A tibble: 142 x 2
#>    hexID pts_list  
#>    <int> <list>    
#>  1    21 <int [12]>
#>  2    22 <int [17]>
#>  3    23 <int [22]>
#>  4    24 <int [10]>
#>  5    25 <int [7]> 
#>  6    26 <int [12]>
#>  7    27 <int [1]> 
#>  8    36 <int [42]>
#>  9    37 <int [42]>
#> 10    38 <int [44]>
#> # i 132 more rows
#> 
#> attr(,"class")
#> [1] "hex_bin_obj"
#> 
#> $model_highd
#> # A tibble: 130 x 8
#>    hexID      x1    x2    x3         x4        x5       x6        x7
#>    <int>   <dbl> <dbl> <dbl>      <dbl>     <dbl>    <dbl>     <dbl>
#>  1    21 -0.992   1.91 1.11  -0.000427   0.000624  0.00749  0.00105 
#>  2    22 -0.906   1.93 1.41  -0.0000183  0.00331  -0.0204  -0.000363
#>  3    23 -0.680   1.93 1.72  -0.000810  -0.00259  -0.00449  0.00153 
#>  4    24 -0.272   1.93 1.96   0.00251    0.00668  -0.0460   0.00128 
#>  5    25  0.0760  1.93 2.00   0.00876    0.00447   0.00851 -0.00195 
#>  6    26  0.461   1.93 1.89  -0.00478    0.00492   0.00835  0.00172 
#>  7    36 -0.985   1.75 0.853 -0.00202    0.000397  0.00331  0.000338
#>  8    37 -0.980   1.66 1.17  -0.000374  -0.00154   0.0165   0.000126
#>  9    38 -0.821   1.64 1.56  -0.000459   0.000538 -0.0123   0.000780
#> 10    39 -0.484   1.68 1.87   0.00313    0.00241   0.00823 -0.00117 
#> # i 120 more rows
#> 
#> $model_2d
#> # A tibble: 130 x 5
#>    hexID   c_x     c_y bin_counts std_counts
#>    <int> <dbl>   <dbl>      <dbl>      <dbl>
#>  1    21 0.358 -0.0171         12     0.16  
#>  2    22 0.441 -0.0171         17     0.227 
#>  3    23 0.524 -0.0171         22     0.293 
#>  4    24 0.608 -0.0171         10     0.133 
#>  5    25 0.691 -0.0171          7     0.0933
#>  6    26 0.774 -0.0171         12     0.16  
#>  7    36 0.316  0.0550         42     0.56  
#>  8    37 0.400  0.0550         42     0.56  
#>  9    38 0.483  0.0550         44     0.587 
#> 10    39 0.566  0.0550         39     0.52  
#> # i 120 more rows
#> 
#> $trimesh_data
#> # A tibble: 322 x 8
#>     from    to x_from  y_from  x_to   y_to from_count to_count
#>    <int> <int>  <dbl>   <dbl> <dbl>  <dbl>      <dbl>    <dbl>
#>  1    16    26 0.275   0.127  0.233 0.199          65       67
#>  2    25    26 0.150   0.199  0.233 0.199          39       67
#>  3    25    37 0.150   0.199  0.191 0.271          39       62
#>  4    36    49 0.108   0.271  0.150 0.343          62       34
#>  5    36    37 0.108   0.271  0.191 0.271          62       62
#>  6     1     7 0.358  -0.0171 0.316 0.0550         12       42
#>  7    48    49 0.0665  0.343  0.150 0.343          45       34
#>  8    37    38 0.191   0.271  0.275 0.271          62       68
#>  9    26    27 0.233   0.199  0.316 0.199          67       56
#> 10    27    38 0.316   0.199  0.275 0.271          56       68
#> # i 312 more rows
\end{verbatim}

@@

Limb muscle cells of mice in \citet{tabula2018} are examined. There are \(1067\) single cells, with \(14997\) gene expressions. Following their pre-processing, tSNE was performed using ten principal components. \ref{fig:nldr-layouts} (b) is the reproduction of the published plot. The question is whether this accurately represents the cluster structure in the data. Our method can help here, and also help to provide a better \(2-\text{D}\) layout, as needed.

\begin{figure}
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/limb-rmse-1} \caption{Assessing which of the 6 NLDR layouts on the limb muscle data  (shown in \\ref{fig:nldr-layouts}) is the better representation using RMSE for varying binwidth ($a_1$). Colour  used for the lines and points in the left plot and in the scatterplots represents NLDR layout (a-f). Layout d is perform well at large binwidth (where the binwidth is not enough to capture the data struture) and poorly as bin width decreases. Layout f is the best choice.}\label{fig:limb-rmse}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/model-limb-1} \caption{Compare the published $2-\text{D}$ layout (\\ref{fig:nldr-layouts} b) and the $2-\text{D}$ layout selected (\\ref{fig:nldr-layouts} f) by RMSE plot (\\ref{fig:limb-rmse}) from the tSNE, UMAP, PHATE, TriMAP, and PaCMAP with different (hyper)parameters. The Limb muscle data ($n =  1067$) has seven close different shaped clusters in $10\text{-}D$.}\label{fig:model-limb}
\end{figure}

>>>>

Non-linear dimension reduction (NLDR) techniques, such as t-distributed stochastic neighbor embedding (tSNE) \citep{laurens2008}, uniform manifold approximation and projection (UMAP) \citep{leland2018}, potential of heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm \citep{moon2019}, large-scale dimensionality reduction Using triplets (TriMAP) \citep{amid2019}, and pairwise controlled manifold approximation (PaCMAP) \citep{yingfan2021}, create wildly different representations depending on the selected method and (hyper)parameter choices. It is difficult to determine whether any of these representations are accurate, which one is the best, or whether they have missed important structures.

@@

@@

Limb muscle cells of mice in \citet{tabula2018} are examined. There are \(1067\) single cells, with \(14997\) gene expressions. Following their pre-processing, tSNE was performed using ten principal components. Figure \ref{fig:limb-rmse} (b) is the reproduction of the published plot. The question is whether this accurately represents the cluster structure in the data. Our method can help here, and also help to provide a better \(2-\text{D}\) layout, as needed.

\begin{figure}
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/limb-rmse-1} \caption{Assessing which of the 6 NLDR layouts on the limb muscle data is the better representation using RMSE for varying binwidth ($a_1$). Colour  used for the lines and points in the left plot and in the scatterplots represents NLDR layout (a-f). Layout d is perform well at large binwidth (where the binwidth is not enough to capture the data struture) and poorly as bin width decreases. Layout f is the best choice.}\label{fig:limb-rmse}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/model-limb-1} \caption{Compare the published $2-\text{D}$ layout (Figure \\ref{fig:limb-rmse} b) and the $2-\text{D}$ layout selected (Figure \\ref{fig:limb-rmse} f) by RMSE plot (Figure \\ref{fig:limb-rmse}) from the tSNE, UMAP, PHATE, TriMAP, and PaCMAP with different (hyper)parameters. The Limb muscle data ($n =  1067$) has seven close different shaped clusters in $10\text{-}D$.}\label{fig:model-limb}
\end{figure}
